---
title: "`coda.base` features' enumeration"
author: "Marc Comas-Cufí"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{coda.base features' enumaration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library(coda.base)
```

# Main functions

* Simulated data set:

```{r}
nP = 6
nO = 8
set.seed(1)
X = matrix(rlnorm(nP * nO), ncol = nP, nrow = nO)
colnames(X) = paste0('P', 1:nP)
rownames(X) = paste0('O', 1:nO)
X
```

####   Descriptive statistics

* Center:

```{r}
center(X)
```

* Variation array:

```{r}
variation_array(X)
```

#### Aitchison distances between compositions

`stats::dist()` is rewritten to include the Aitchison distance between compositions:

```{r}
dist(X, method = 'aitchison')
```


#### Log-ratio coordinates. `coordinates()`

```{r}
coordinates(X)
```

By default, `coda.base` uses the isometric log-ratio coordinates defined in Egozcue et al. 2003.

-   `coordinates(X, 'ilr')`: isometric log-ratio coordinates (Egozcue et al. 2003, defaults)
-   `coordinates(X, 'alr')`: additive log-ratio coordinates
-   `coordinates(X, 'clr')`: centered log-ratio coordinates
-   `coordinates(X, 'pw')`: pairwise log-ratio coordinates
-   `coordinates(X, 'pc')`: principal component coordinates
-   `coordinates(X, 'pb')`: principal balance coordinates
-   `coordinates(X, 'cdp')`: balanced isometric log-ratio coordinates

`coordinates(X, B)` accepts a log-contrast matrix $B$ to build the log-ratio coordinates. Different log-contrast matrices $B$ can be constructed (following section).

#### Functions to build log-contrast matrices

```{r}
ilr_basis(nP)
```

```{r}
all.equal(as.numeric(coordinates(X, 'ilr')),
          as.numeric(log(X) %*% ilr_basis(nP)))
```

Log-ratio matrix transformations:

- `ilr_basis(nP)` (Egozcue et al. 2003, defaults)
- `ilr_basis(nP, type = 'pivot')` to pivot log-ratio coordinates
- `ilr_basis(nP, type = 'cdp')` to balanced log-ratio coordinates (CoDaPack's default)
- `alr_basis(nP)` to additive-log ratio coordinates. Numerator order and denominator can be modified. For example, `alr_basis(nP, denominator = 1, numerator = nP:2)`.
- `clr_basis(nP)` to centered log-ratio coordinates.
- `pc_basis(X)` to principal components log-ratio coordinates.
- `cc_basis(X, X2)` to canonical correlations log-ratio coordinates.
- `pb_basis(X, method = 'exact')` to principal balances using the exact algorithm.
- `pb_basis(X, method = 'constrained')` to principal balances using pca constrained algorithm.
- `pb_basis(X, method = 'cluster')` to principal balances obtained using parts clustering algorithm.
- `pairwise_basis(nP)` to pairwise log-ratio coordinates.
- `sbp_basis(b0 = b1~b2, b1 = P1~P2+P3, b2 = P4~P5+P6, data=X)`   

