---
title: "Introducing `coda.base`"
author: "Marc Comas-Cufí"
date: "2019-09-26"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
set.seed(1)
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(echo = TRUE, comment = "#>  ", fig.retina = 4)
```

---
class: inverse, center, middle

# Getting Started

---
class: middle

# A composition

```{r, message=FALSE}
library(coda.base)
X = as.data.frame(exp(matrix(rnorm(5*6), ncol = 5)))
X
```

---
class: inverse, center, middle

# Clasical coordinates

---

### ALR

```{r}
coordinates(X, basis = 'alr')
```

---

### CLR

```{r}
coordinates(X, basis = 'clr')
```

---

### ILR (default)

```{r}
coordinates(X, basis = 'ilr')
all.equal(coordinates(X, basis = 'ilr'),
          coordinates(X))
```

---
class: inverse, center, middle

# Performance

```{r, include=FALSE}
library(ggplot2)
theme_set(theme_minimal())
library(microbenchmark)
load('bench.RData')
```

---

### ALR performance

```{r, echo=FALSE, fig.width=6, fig.height=4, message=FALSE, fig.align='center', out.width= "80%"}
autoplot(alr_bench) + labs(title = 'ALR coordinates performance')
```

---

### CLR performance

```{r, echo=FALSE, fig.width=6, fig.height=4, message=FALSE, fig.align='center', out.width= "80%"}
autoplot(clr_bench) + labs(title = 'CLR coordinates performance')
```

---

### ILR performance

```{r, echo=FALSE, fig.width=6, fig.height=4, message=FALSE, fig.align='center', out.width= "80%"}
autoplot(ilr_bench) + labs(title = 'ILR coordinates performance')
```

---

### Matrix based tranformation performance

```{r, echo=FALSE, fig.width=6, fig.height=4, message=FALSE, fig.align='center', out.width= "80%"}
autoplot(matrix_bench) + labs(title = 'Coordinates performance')
```

---
class: inverse, center, middle

# Coordinates using basis matrix

---

### ALR matrix (1)

```{r}
B.alr = alr_basis(5)
B.alr
coordinates(X, B.alr)
```

---

### ALR matrix (2)

```{r}
B.alr = alr_basis(5, denominator = 3, numerator = c(2,1,5,4))
B.alr
coordinates(X, B.alr)
```

---

### ILR matrix

```{r}
ilr_basis(5, type = 'default')
```

---

### ILR matrix (pivot)

```{r}
ilr_basis(5, type = 'pivot')
```

---

### ILR matrix (CoDaPack)

```{r}
ilr_basis(5, type = 'cdp')
```

---
class: inverse, center, middle

# Balances

---

### Sequential binary partition

```{r}
bp = matrix(c(+1,+1,-1,-1, 0,
              +1,+1,+1,+1,-1,
              +1,-1, 0, 0, 0,
               0, 0,+1,-1, 0), ncol = 4)
sbp_basis(bp)
```

---

### Sequential binary partition

```{r}
sbp_basis(V1+V2~V3+V4,
          V1+V2+V3+V4~V5,
          V1~V2,
          V3~V4, data = X)
```

---

### Sequential binary partition

```{r}
sbp_basis(b1 = b3~b4,
          b2 = b1~V5,
          b3 = V1~V2,
          b4 = V3~V4, data = X)
```

---
class: inverse, center, middle

# Other coordinates

---

# Principal components

```{r}
H.pc = coordinates(X, basis = 'pc')
attr(H.pc, 'basis')
apply(H.pc, 2, var)
```

---

# Principal balances

```{r}
H.pb = coordinates(X, basis = 'pb')
attr(H.pb, 'basis')
apply(H.pb, 2, var)
```

* Variations: Ward method, local search optimisation, simulated annealing, tabu search.
* Other optimality measures: discriminant balances, canonical balances, ...

---
class: inverse, center, middle

# Obtaining `coda.base`

---

# Installation

### From CRAN

```{r, eval=FALSE}
install.packages()
```

### From Github

```{r, eval=FALSE}
# install.packages('remotes')
remotes::install_github('mcomas/coda.base')
```
